---
title: "Bellabeat Capstone Project"
author: "Salay"
date: '2022-06-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

This is the Case Study Project (Capstone Project) which is the last part of the [Google Data Analytics Professional Certificate](https://www.coursera.org/professional-certificates/google-data-analytics)


**The key business task is to reveal opportunities for growth in the smart health device market.**

The data analyzed is from Kaggle, [FitBit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit).


# Data Processing  

This section will go through the necessary data processing on FitBit datasets.    
  
First of we will load the necessary libraries.  
```{r}
library(tidyverse)
library(lubridate)

```


Now, let us load the data sets that we are going to use.

```{r}
daily_calories <- read.csv("dailyCalories_merged.csv")
daily_sleep <- read.csv("sleepDay_merged.csv")
daily_activity <- read.csv("dailyActivity_merged.csv")

```
### We will start with the "daily_calories" data frame.

```{r}
glimpse(daily_calories)

```
Checking column types and changing to appropriate ones where necessary.

```{r}
daily_calories$ActivityDay <- mdy(daily_calories$ActivityDay)

#Check if character type changed to appropriate
typeof(daily_calories$ActivityDay)

```
Checking for duplicates

```{r}
daily_calories <- daily_calories %>% 
  distinct()

```
Data processing for this data frame is complete
```{r}
glimpse(daily_calories)
```

### Now the second "daily_sleep" data frame.

```{r}
glimpse(daily_sleep)
```
Checking column types and changing to appropriate ones where necessary.

```{r}
daily_sleep$SleepDay <- mdy_hms(daily_sleep$SleepDay)

#Check if character type changed to appropriate
typeof(daily_sleep$SleepDay)

```
Checking for duplicates

```{r}
daily_sleep <- daily_sleep %>%
  distinct()

```
Data processing for this data frame is complete
```{r}
glimpse(daily_sleep)
```
### Now the third "daily_steps" data frame.

```{r}
glimpse(daily_activity)
```

For my analysis, I only requires the data related to users' steps.
```{r}
daily_steps <- daily_activity %>%
  select(Id,ActivityDate,TotalSteps)

```


Checking column types and changing to appropriate ones where necessary.

```{r}
daily_steps$ActivityDate <- mdy(daily_steps$ActivityDate)

#Check if character type changed to appropriate
typeof(daily_steps$ActivityDate)

```
Checking for duplicates

```{r}
daily_steps <- daily_steps %>%
  distinct()

```
Data processing for this data frame is complete
```{r}
glimpse(daily_steps)
```

# Data Analysis and Visualization


## Analysis of users daily calorie expenditure

Basic descriptive statistics on the data by each user to see data consistency
```{r}
cal_expenditure_info <- daily_calories %>%
  group_by(Id) %>% drop_na() %>%
  summarise(mean = mean(Calories), stdev = sd(Calories),
            coef_var = sd(Calories)/mean(Calories))
head(cal_expenditure_info)
```
Plotting each users; calorie expenditure over time
```{r}
daily_calories %>%
  drop_na(Calories) %>%
  ggplot(aes(x=ActivityDay,y=Calories,group=1)) +
  geom_line()+
  facet_wrap(vars(Id))+
  labs(title = "Calorie Burn by User Over Time")+
  theme(plot.title = element_text(hjust = 0.5))

```

## Analysis of users' sleep insufficiency

Finding the mean sleep time of each user
```{r}
daily_sleep_summary <- daily_sleep %>% 
  group_by(Id) %>% 
  summarise(mean = mean(TotalMinutesAsleep))

glimpse(daily_sleep_summary)

```

Checking if there zero values in the result.
```{r}
all(daily_sleep_summary$mean == 0)
```

For the purposes of the analysis, removed mean sleeping hour less than 4.Since
probably this was due to incorrect logging of sleep data.

```{r}
daily_sleep_summary <-  subset(daily_sleep_summary, mean >= 240)

```

Defining necessary variables
```{r}
#Finding the remaining users
users <- n_distinct(daily_sleep_summary$Id)

#The recommended 8 hour sleep by experts
sleep_min <- 480

```

Plotting the users mean sleeping hours to see if they are getting enough sleep.
```{r}
daily_sleep_summary %>%
  ggplot(aes(x=1:users, y=mean)) +
  geom_area()+
  scale_y_continuous(n.breaks = 10, expand = c(0, 0), limits = c(0, NA))+
  scale_x_continuous(n.breaks = 21,expand = c(0, 0), limits = c(1, NA))+
  geom_text(aes(users -3, sleep_min, label = "Recommended 480", vjust = -0.5))+
  geom_hline(yintercept = sleep_min, color = "red", size = 2) +
  labs(title = "Mean Sleep time by user",y = "Sleep time (minutes)",x = "User")+
  theme(plot.title = element_text(hjust = 0.5))

```

## Analysis of users' activeness

Finding total sum of steps taken by each users
```{r}
steps_sum_user <- daily_steps %>% 
  group_by(Id) %>% 
  summarise(sum_of_steps = sum(TotalSteps))
```

Checking if there any users who have zero total steps
```{r}
all(steps_sum_user$sum_of_steps == 0)
```

Finding most and least 5 users with total sum of steps

```{r}
#finding 5 with the most amount of steps
most_active_users <- head(arrange(steps_sum_user, -sum_of_steps),5) %>%
  select(Id)

#finding 5 with the least amount of steps
least_active_users <- head(arrange(steps_sum_user, sum_of_steps),5) %>%
  select(Id)

```

Creating a data frame to analyze the top 5 and bottom 5 users by activeness

```{r}

daily_steps_mostactive <-inner_join(daily_steps,most_active_users, by ="Id")
daily_steps_leastactive <-inner_join(daily_steps,least_active_users, by ="Id")


```

Classification of the sum of steps by activeness

```{r}
activity_classfication_most_active <- daily_steps_mostactive %>% 
  group_by(Id) %>% 
  summarise("sedentary" = sum(TotalSteps<5000),
            "low_active" = sum(TotalSteps>=5000 & TotalSteps<7500),
            "active" = sum(TotalSteps>=7500 & TotalSteps<10000),
            "very_active" = sum(TotalSteps>=10000))

activity_classfication_least_active <- daily_steps_leastactive %>% 
  group_by(Id) %>% 
  summarise("sedentary" = sum(TotalSteps<5000),
            "low_active" = sum(TotalSteps>=5000 & TotalSteps<7500),
            "active" = sum(TotalSteps>=7500 & TotalSteps<10000),
            "very_active" = sum(TotalSteps>=10000))

head(activity_classfication_least_active)
head(activity_classfication_most_active)
```

Finding the mean of the top 5 and bottom for activeness classifcation
```{r}
#converting Id to string to avoid calculation
activity_classfication_most_active$Id<-
  as.character(activity_classfication_most_active$Id)
activity_classfication_least_active$Id <- as.character(activity_classfication_least_active$Id)

```

```{r}
#finding average of least active by class
least_active <- colMeans(activity_classfication_least_active[
  sapply(activity_classfication_least_active,is.numeric)])

activeness_summary <- data.frame(least_active)
```


```{r}
#finding average of most active by class
most_active <- 
  colMeans(activity_classfication_most_active[sapply(activity_classfication_most_active, is.numeric)])

#adding the result to the data frame
activeness_summary$most_active <- most_active

activeness_summary <- activeness_summary[!(row.names(activeness_summary) %in% "Id"),]

```
Let us look at the end result of this.

```{r}
head(activeness_summary)
```

Visualizing the activeness for most and least active based on the classification

```{r}
activeness_summary %>%
  rownames_to_column("ID") %>%
  pivot_longer(!ID) %>%
  ggplot(aes(x= fct_inorder(ID), y = value, fill = name))+
           geom_bar(position="fill", stat="identity")+
  labs(title = "Most Active vs Least Active Users by Activity Class",y = "Total",x = "Activineness")+
  theme(plot.title = element_text(hjust = 0.5))

```

Plot to see if there is any change in activeness over time in both categories

Summarizing the most and least active users over time

```{r}
leastactive_summary <- daily_steps_leastactive %>%
  group_by(ActivityDate) %>%
  summarise(mean_leastactive = mean(TotalSteps))
  
mostactive_summary <- daily_steps_mostactive %>%
  group_by(ActivityDate) %>%
  summarise(mean_mostactive = mean(TotalSteps))

activeness_overtime <- inner_join(leastactive_summary,mostactive_summary, by = "ActivityDate")

glimpse(activeness_overtime)

```

Visualization

```{r}
activeness_overtime %>%
  ggplot(aes(x=ActivityDate))+
  geom_line(aes(y = mean_mostactive, color = "Most Active"))+
  geom_line(aes(y = mean_leastactive, color="Least Active"), linetype="twodash")+
  scale_color_manual(values = c("Most Active" = "darkred", 
                                "Least Active" = "steelblue"))+
  labs(title = "Users' Activeness Steps Over Time",y = "Steps",x = "Date")+
  theme(plot.title = element_text(hjust = 0.5))
```



